---
title: "<h1 style='text-align:center;'> Sample size calculation </h1>"
author: "Hugo Marthinet"
date: "`r Sys.Date()`"
output:
  knitr:::html_vignette:
  toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Sample size calculation}
  %\usepackage[utf8]{inputenc}
---

\setmainfont{Calibri Light}

```{r, echo = FALSE, message = FALSE}

# If you want to create a PDF document paste the following after line 9 above:
#   pdf_document:
#     toc: true
#     highlight: tango
#     number_sections: no
#     latex_engine: xelatex    
# header-includes: 
#    - \usepackage{fontspec}

knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(sampleSize)
```

This packages provides sample size calculation functions for various designs not provided in **pwr* and **webPower** packages.
Every sample size calculation requires some kind of estimation. One must, wether it is from the literature, previous studies or field experts, provide some estimation of 
parameters of interest. This parameters vary depending on the study design and can be quite hard to determine in some cases.
Based on this assumption, sample size calculation in itself is only an estimation ; this should not be neglected when it comes to the analysis. 
Loose parameters estimation will lead to weak analysis.


## Descriptive study

One usually only use sample size determination in a descriptive study context whenever no further statistical analysis are planned, or when no assumptions are made over the study whatsoever.
In this case, one should assume the hypothesis that every answer to the survey will be evenly distributed among the possible answers.
This function thus should only be used as a last resort if no other calculation fit the study design. In this case, the only parameter to estimate in the size of the studied population.

```{r, message = FALSE}
sampleSize::SSdescriptive(70000000)

sampleSize::SSdescriptive(70000000, e = 0.01)
```

When the size of error margin is 5%, the sample size is soft caped at 400, even when considering the whole french population of roughly 70 million people.
However, this quickly skyrocket whenever the error margin is reduced.

## Cohort study

Using the example of a disease affecting 0.2% of the exposes population, and 0.6% of the unexposed population (risk factor of 3 in the unexposed group).
If the cohort is planned to last 5 years and 30% of the patients inclued are to receive the treatment, the sample size calcultaion is as follow :

```{r, message = FALSE}
# Using the risk.factor parameter
SScohort(incidence.exp = 2/1000, risk.factor = 3,
         prop.treatment = 0.3, time = 5)

# Using the incidence.nexp parameter
SScohort(incidence.exp = 2/1000, incidence.nexp = 6/1000, 
         prop.treatment = 0.3, time = 5)
```

At least 1743 patients/year should be included.

## Means, proportions and others

All of the study below roughly require the same parameters estimation of the study population : means, proportions, prevalence... in the different groups.
This is always paired with **alpha**, **power** and an **alternative hypothesis**. Examples below are from real case study :

```{r, message = FALSE}
# Proportion study
# 20% prevalence in treated group versus 22% in the untreated group
# Prevalence are estimated from a pre-study
SSproportion(0.20, 0.22)

# Mean study 
# Calories consumption : average of 2350 in males, 1850 in females
# Means and standard deviation are estimated from previous studies
SSmean(m1 = 2350, m2 = 1850, sd1 = 250, sd2 = 400)

# Sensibility study
# Sample size requirment to proove the new screening test is more efficatious then the previous one
# Known sensibility and psecificity of the previous screening test of 90%, prevalence estimated by experts of 55%
SSsensibility(sensi = 0.9, speci = 0.9, prevalence = 0.55)
```

## Linear and logistic regressions

Functions is this section do exist in other packages, but the version provided here are more user friendly. Linear and logistic regression study sample size calculation are quite 
straight forward and require few estimation. This is, in a linear regression, only the R estimation, and prevalence in both group fro the logsitic regression.

```{r, message = FALSE}
# Linear regression
# Data from,previous study
X <- c(46.8, 48.7, 48.4, 53.7, 56.7)
Y <- c(14.6, 19.6, 18.6, 25.5, 20.4)

mod <- lm(Y ~ X)
SSlinearReg(df = summary(mod)$fstatistic[2],
            R = sqrt(summary(mod)$adj.r.squared))
SSlinearReg(mod)



# Logistic regression
# Expert estimation
SSlogisticReg(0.75, 0.85, family = "normal")
```
